Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-04 paper.tex
restructured

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disk such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depends on the height from the 
  disk midplane, which is generally expected in  
  astrophysical disks. However, the VSI requires a short thermal
  timescale in stably stratified disks. %which is usually assumed 
  A quantitative examination of this thermodynamic 
  requirement is thus necessary to assess the viability of
  the VSI in realistic disks. To this end,  
  we generalize the linear, axisymmetric VSI in radially local but
  vertically global models of accretion disks to include a thermal
  relaxation term that restores the disk  
  temperature $T$ over a characteristic timescale
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response  of the disk by
  varying $\beta$. We find increasing $\beta$ even slightly away from
  zero can dramatically reduce VSI growth rates. % , since this generally
  % introduces a vertical entropy gradient that stabilizes vertical
  % motions as needed for the VSI.
  Our  linear calculations are consistent with previous hydrodynamic
  simulations. For vertically isothermal disks, we quantify the
  requirement for `fast' thermal relaxation as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  ($r$
  being the cylindrical radius from the central object), and 
  $\gamma$ is the adiabatic index. We apply our linear calculations 
  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
  estimate for the thermal timescale $\beta$ in the MMSN based on dust
  opacity. We find the MMSN is unstable to the VSI at tens of AU with
  a characteristic growth times of $\sim 30$ local orbital periods. 
\end{abstract}

>>>>

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disk such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depends on the height from the 
  disk midplane, which is generally expected in  
  astrophysical disks. However, the VSI requires a short thermal
  timescale in stably stratified disks. %which is usually assumed 
  A quantitative examination of this thermodynamic 
  requirement is thus necessary to assess the viability of
  the VSI in realistic disks. To this end,  
  we generalize the linear, axisymmetric VSI in radially local but
  vertically global models of accretion disks to include a thermal
  relaxation term that restores the disk  
  temperature $T$ over a characteristic timescale
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response  of the disk by
  varying $\beta$. We find increasing $\beta$ even slightly away from
  zero can dramatically reduce VSI growth rates. % , since this generally
  % introduces a vertical entropy gradient that stabilizes vertical
  % motions as needed for the VSI.
  Our  linear calculations are consistent with previous hydrodynamic
  simulations. For vertically isothermal disks, we quantify the
  requirement for `fast' thermal relaxation as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  ($r$
  being the cylindrical radius from the central object), and 
  $\gamma$ is the adiabatic index. We apply our linear calculations 
  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
  estimate for the thermal timescale $\beta$ in the MMSN based on dust
  opacity. We find the MMSN is unstable to the VSI at tens of AU with 
  a characteristic growth time of $\sim 30$ local orbital
  periods, and the dominant disturbance in the MMSN is the `corrugation' mode
  which casues the disk column to oscillate vertically. 
\end{abstract}
